\section{Discussion}
\label{sec:discussion}

% (~2 pages)

\paragraph{Instruction-following as a trainable, content-dependent property.}
The property we measure is descriptive: a model's susceptibility to in-context pressure, the same mechanism exploited by many-shot jailbreaking \citep{anil_many-shot_2024} and Crescendo-style escalation \citep{russinovich2025great}, and one that is largely decorrelated from capability benchmarks---a dimension that leaderboards do not capture.
As previously found by \citet{pyatkin_generalizing_2025}, instruction-following robustness is not cleanly correlated with general capability. This also applies to our experimental setting, in which scale and benchmark performance are poor predictors of resistance to induction pressure. 
The OLMo training-stage comparison suggests where robustness might be installed: DPO (preference training) produces the largest gain, compared to SFT and RLVR and in contrast to the finding from \citet{pyatkin_generalizing_2025}. Advanced instruction tuning, like the one performed on Llama 3.3 vs.\ 3.1 \citep{llama33_2024}, provides further but more modest improvement. 
Robustness also interacts with instruction content. Our results suggest that instruction-following is not a content-neutral capability. The model's value priors, instilled during post-training, modulate how strongly it adheres to instructions in our conflictual setting. This has direct implications for alignment: models are more reliable when instructions happen to agree with their trained preferences.


\paragraph{Output diversity as a source of robustness.}
In our setting, task-based conditions lead to higher instruction-following rates than fixed-output conditions, despite measuring the same underlying conflict. A targeted followup (Appendix~\ref{app:followup}) shows that the gap is not explained by task difficulty or semantic engagement, but is sensitive to output diversity, which has a positive effect on robustness. Mechanistically, this gap reflects a competition between low-level token-copying circuits \citep{yona2025interpreting} and high-level latent task representations \citep{toddFunctionVectorsLarge2024}. Single-token outputs are highly susceptible to induction head toxicity, where specialized attention heads for verbatim matching dominate the output logits and suppress other computations \citep{yinattention}. Free-text generation and the associated ''Function vectors'' might be fundamentally more robust to contextual pressure and repetition. %This asymmetry has a geometric counterpart: \citet{hosseini_context_2026} show that continual prediction settings such as natural language generation induce representational straightening---a linearization of neural trajectories in deep layers that facilitates stable next-token prediction---whereas this stabilizing structure vanishes in structured prediction settings.
Another hypothesis is that the varied token distribution of natural language may further reinforce the stability of the assistant persona \citep{luAssistantAxisSituating2026}, allowing the model to remain in ``instruction-following mode'' longer when its outputs resemble those of a helpful assistant. This has a practical implication: robustness measures derived from fixed-output conditions may underestimate a model's resistance to induction in realistic deployments, where responses are naturally varied. That said, existing work shows that models can still be pushed towards misalignment by malicious in-context examples \citep{afonin_emergent_2026}.

\paragraph{Limitations.}
% Llama 3's behavior, with its surprising robustness that does not degrade even at high Ns remains unexplained. It is not predicted by scale, architecture, or benchmark performance.
Our work presents several limitations. We recognize that the experimental setting is controlled but artificial. Real-world instruction override occurs through more naturalistic mechanisms; whether the same factors modulate robustness in deployed settings is an open empirical question.
Greedy decoding at T=0 captures modal behavior but not the full output distribution. The $T{=}1$ comparison (Appendix~\ref{app:temperature}) partially addresses this, but shows large model-specific variance.


% \paragraph{Future work.}
% % Linear probe analysis tracking the instruction/induction axis through
% % internal representations (cf. Lampinen et al.).
% % Endorsement protocol: prefilling responses and testing whether models
% % endorse or disavow them.
% % Extension to naturalistic settings (multi-turn conversations where
% % the induction pressure is implicit rather than explicit).

